---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tRexCAP

<!-- badges: start -->
<!-- badges: end -->

tRex-CAP uses a divide and conquer strategy and the tREX algorithm as the building block for constructing whole genome three-dimensional structures.

## Installation

You can install tRex-CAP package directly from github by using your favorite installer (remotes/devtools/pak):

```{r}
#| eval: false
remotes::install_github("osu-stat-gen/tREX-cap@dev")
```

## Example

To run the cut portion of the algorithm, call `Cut` function directly


```{r example, eval=FALSE}
library(tRexCAP)
data(sim_hic)
data(sim_bias)
CutRes = Cut(contact = sim_hic, bias = sim_bias, block_size = 12, noverlap = 1, CPU = 5)
```

To run the paste algorithm, use `Paste` function.

```{r, eval=FALSE}
PasteRes = Paste(contact = sim_hic, cutresult = CutRes, CPU = 5)
```

It is also possible to run both sequentially by calling `CutAndPaste` function.

```{r, eval=FALSE}
CAPRes = CutAndPaste(contact = sim_hic, block_size = 12, noverlap = 1, CPU = 5)
```