---
title: "tRexCAP: A Divide and Conquer Strategy for Recapitulating Whole Genome 3D Structure Using Hi-C Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tRexCAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
The spatial organization of the genome is a critical determinant of cellular function, yet its complex architecture cannot be fully captured through linear sequencing alone. Hi-C technology overcomes this limitation by capturing chromosomal conformation information between loci. Translating the resulting 2D proximity matrices into physical 3D models enables the identification of spatial "neighborhoods," revealing how non-coding genetic variants exert regulatory influence over vast genomic distances.

tRex-CAP (Cut and Paste) is designed to reconstruct whole-genome three-dimensional structures from Hi-C data. The method employs a divide-and-conquer strategy to manage the complexity of genomic data: first, each chromosome is partitioned into $n$ overlapping blocks; second, the tRex algorithm reconstitutes the 3D structure of each block independently; finally, these local structures are "pasted" together to form a global model while preserving the structural rigidity of the individual blocks.

In this pacakge, we provide the following main functions:

* Cut
* Paste
* CutAndPaste
* DrawStucture

We illustrate the usage of these functions in the following sections.

# Example Data

An example contact matrix of n=1000 loci simulated with Poisson model with random effects can be loaded by calling `data(example_data)`. The matrix should be \eqn{n \times n} and symmetric, where the (i, j) entry of the matrix is the count of interactions between loci i and j. The diagnoal elements of the matrix should be 0.


```{r setup}
library(tRexCAP)
data(example_data)
dim(example_data)
```

The optional bias matrix should be of dimension \eqn{n \times 3}. The 3 columns describe effective fragment information, GC content, and mappability for each locus, respectively. 

```{r bias}
data(example_bias)
head(example_bias)
```


# tRex-CAP
## Cut 

For simplicity, we use the first 100 loci in our example

Once the data is loaded, Cut-and-Paste algorithm can run separately by calling `Cut` and `Paste` in two separate steps.

```{r, eval=FALSE}


example_data_short <- example_data[1:100, 1:100]
example_data_short_bias <- example_bias[1:100:,]
control_trex <- list(
  iter = c(500, 500, 500, 500, 500),
  block_size = c(10, 10, 10, 10, 10),
  max_cycle = 1
)

# these examples run on a 100x100 contact matrix 

# should take around 21 seconds to run this cut with 1 CPU on 100 x 100 contact
CutRes = Cut(contact = example_data_short, 
            block_size = 25, 
            noverlap = 1, 
            CPU = 1,
            mcmc = 500,
            burn = 500,
            thinning = 1,
            target.accept = 0.95,
            tuning_control = control_trex)

PasteRes = Paste(contact = example_data_short,  
                cutresult = CutRes, 
                CPU = 1)
```

Alternatively, it can be run in a single step by calling `CutAndPaste`
```{r, eval=FALSE}
CAPRes = CutAndPaste(contact = sim_hic, block_size = 12, noverlap = 1, CPU = 5)
```

## Paste

## CutAndPaste
## Results 

## Visualization

# Computation Cost

# Example 

# Versions

# Reference