---
title: "tRexCAP: A Divide and Conquer Strategy for Recapitulating Whole Genome 3D Structure Using Hi-C Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tRexCAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
The spatial organization of the genome is a critical determinant of cellular function, yet its complex architecture cannot be fully captured through linear sequencing alone. Hi-C technology overcomes this limitation by capturing chromosomal conformation information between loci. Translating the resulting 2D proximity matrices into physical 3D models enables the identification of spatial "neighborhoods," revealing how non-coding genetic variants exert regulatory influence over vast genomic distances.

tRex-CAP (Cut and Paste) is designed to reconstruct whole-genome three-dimensional structures from Hi-C data. The method employs a divide-and-conquer strategy to manage the complexity of genomic data: first, each chromosome is partitioned into $n$ overlapping blocks; second, the tRex algorithm reconstitutes the 3D structure of each block independently; finally, these local structures are "pasted" together to form a global model while preserving the structural rigidity of the individual blocks.

In this pacakge, we provide the following main functions:

* Cut
* Paste
* CutAndPaste
* DrawStucture

We illustrate the usage of these functions in the following sections.

# tRex-CAP

tRex-CAP algorithm consists of two steps. The first step is `Cut` where the whole contact matrix is partitioned into $n$ blocks (with overlaps), and tRex algorithm is run on each block to obtain their own structure The second step is `Paste` where we merge all blocks into one structure, using isometric transformations to preserve the structure of each block, while matching them at the overlapping loci.

## Cut

`Cut` function requries the following arguments:
* Contact: The matrix should be \eqn{n \times n} and symmetric, where the (i, j) entry of the matrix is the count of interactions between loci i and j. The diagnoal elements of the matrix should be 0. An example contact matrix of n=1000 loci simulated with Poisson model with random effects can be loaded by calling `data(example_data)`.

```{r setup}
library(tRexCAP)
data(example_data)
dim(example_data)
example_data[1:5,1:5]
```

* block_size: The size of each block.
* noverlap: Number of loci that overlap between neighboring blocks.

Other other arguments are optional. 

* bias: The optional bias matrix should be of dimension \eqn{n \times 3}. The 3 columns describe effective fragment information, GC content, and mappability for each locus, respectively. 

```{r bias}
data(example_bias)
head(example_bias)
```

* breaks: A vector that describes the indices of pre-specified cut points. If unspecified, equal block size cuts will be used.
* CPU: Integer specifying the number of cores for parallel MCMC execution. 
* save_mcmc: Whether to save the posterior samples generated during the MCMC step. 
* mcmc, burn, Number of HMC iterations and burn-ins, respectively.
* thinning: Thinning rate. For example, a thinning rate of 10 means every 10th HMC iteration is saved as posterior samples.
* jump: The initial standard deviations for parameters updated via Metropolis-Hastings.
* leapfrog.L, leapfrog.e The initial number of leapfrog steps and intial epsilon  for HMC.
* target.accept The target acceptance rate. 
* tuning_control Tuning control for HMC. See below for details.
* method.type One of tRex, tPAM and bn.

## Tuning Control
We do a five phase tuning. Phase 1 adapts the number of steps (L) and the step size (epsilon), phase 2 runs the chain, phase 3 adapts the step size (epsilon), phase 4 runs the chain, and phase 5 adapts the Metropolis-Hasting jump parameters.

The tuning control is a list of iter, block_size and max_cycle. iter controls the number of iterations for each phase. block_size here controls how often acceptances rates are reported and hyperparameter adjustments occur according to parameter acceptance rates. max_cycle determines how many times we go through the phases. Between cycles, the structure is reset but hyperparameter settings are retained. By default, the settings for tuning_control is retained. Here is an example of tuning control we used for the sample data.

```{r controls}
control_trex <- list(
  iter = c(500, 500, 500, 500, 500),
  block_size = c(10, 10, 10, 10, 10),
  max_cycle = 1
)
```

### Output

`Cut` function returns the sturcture of each block and other parameters intended for `Paste`.

## Paste

`Pates` function requires the following arguments:


* contact: The same contact matrix used in `Cut`.
* cutresult: The result from a Cut() call.
* CPU:  Integer specifying the number of cores for parallel MCMC execution.

### Output
`Paste` returns an \eqn{n \times 3} matrix of the estimated coordinates.

### Example
For simplicity, we use the first 100 loci in our example with no bias matrix.

```{r, results = 'hide', message = FALSE}
example_data_short <- example_data[1:100, 1:100]
CutRes <- Cut(
            contact = example_data_short, 
            block_size = 25, 
            noverlap = 1, 
            CPU = 1,
            mcmc = 1000,
            burn = 1000,
            thinning = 1,
            target.accept = 0.95,
            tuning_control = control_trex)
```

On the 3303MHz AMD EPYC 7R32 CPU, without parallelization it took 14.7 seconds to execute the above command.

To paste the results, 
```{r}
PasteRes <- Paste(
              contact = example_data_short,  
              cutresult = CutRes, 
              CPU = 1)
```

On the 3303MHz AMD EPYC 7R32 CPU, without parallelization it took 23.8 seconds to execute the above command.

## CutAndPaste
We also wrapped `Cut` and `Paste` into a single call, `CutAndPaste`. The arguments for `CutAndPaste` is the same as `Cut`. It runs `Cut` and then `Paste` sequentially, and returns the coordinates matrix.

```{r, eval=FALSE}
CAPRes <- CutAndPaste(
            contact = sim_hic,
            block_size = 12,
            noverlap = 1,
            CPU = 5)
```

# Visualization
Once we obtain the coordinate matrix, we can plot the 3D structure with `draw_struct`

```{r}
draw_struct(PasteRes)
```

# Simulation
Here we attach our script for data simulation.
## Simulation Function
For a known structure S, we simulate the data based on either Poisson or negative binomial distrubtion. Possible parameters includes pi for zero inflation percentage, p for overdispersion paramter for negative binomial distribution, and random for the presence of random effect. 
```{r}
gen_contact <- function(S, family = "poisson", pi = 0, p = 0.75, random=FALSE) {
    # S matrix: coordiantes of the loci
    # family: poisson or NB
    # pi: zero inflation percentage: 0, 0.05, 0.15, 0.25, or 0.5
    # p overdispersion factor for NB: 0.5, 0.75, or 0.9
    # family='poisson' or 'nb'
    ns <- nrow(S)
    nI <- ns * (ns - 1) / 2
    alpha0 <- 4  # original is 2
    alpha1 <- -1.2
    dist_S <- dist(S)
    c1 <- matrix(nrow = ns, ncol = ns, 0)
    # dist is the lower triangle matrix
    # mu vector
    #no random component
    mu <- exp(alpha0 + alpha1 * log(dist_S))

    if (random) {
        NoiseU2 <- rnorm(nI, sd = sqrt(0.03))
        X <- rnorm(ns, sd = sqrt(0.1))
        temp <- matrix(rep(X, ns), ncol = ns, byrow = T)
        X_mat <- as.vector((temp + t(temp))[lower.tri(temp)])
        mu <- mu + X_mat + NoiseU2
        mu[mu < 1e-6] <- 1e-6
    }

    if (family == "poisson") {
        counts <- rzipois(nI, lambda = mu, pstr0 = pi)
    } else {
        counts <- rzinegbin(nI, size = mu * p / (1 - p), munb = mu, pstr0 = pi)
    }
    c1[lower.tri(c1)] <- counts
    return(c1)
}
```

## Simulation
For the structure we used the first 500 and the last 500 loci from Lieberman-Aiden et al., 2009. The estimated whole genome structure is available in `data(lieberman)`.

```{r, eval = FALSE}
require(VGAM)
require(reticulate)
require(splines)
# turn off scientific notation
options(scipen = 999)
data(lieberman)
N <- nrow(lieberman)
# in our simulation we take the first 500 coordinates and last 500 coordinates of the leibermanS
S <- lieberman[c(1:500, (N - 499):N),]

# settings
family <- "poisson" # or nb for negative binomial 
pi <- 0 # zero-inflation parameter: options used in manuscript: 0, 0.05, 0.15, 0.25, 0.5
p <- 0.75 # overdispersion parameter: options used in manuscript: 0.5, 0.75, 0.9
random <- TRUE # random effect: if TRUE use random effect terms if FALSE do not use
nreps <- 1  # we set nreps to 100 in tRex-cap manuscript
seed = 2  # seed used in manuscript
cat("Simulating ", family, " with pi=", pi, ", p=", p, " random: ", random, " nreps = ", nreps, "\n", sep="")
set.seed(seed)

# build an informative output directory name
if (family == "poisson") {
  outdir <- paste0(
    "../siml_data/",
    family,
    "_pi_", pi,
    ifelse(random, "_ran", "_noran")
  )
} else {
  outdir <- paste0(
    "../siml_data/",
    family,
    "_pi_", pi,
    "_p_", p,
    ifelse(random, "_ran", "_noran")
  )
}

# create it if it doesn't exist
if (!dir.exists(outdir)) {
  dir.create(outdir, recursive = TRUE)
}


for (i in 1:nreps) {
    contact <- gen_contact(S, family = family, pi = pi, p = p, random = random)
    contact <- contact + t(contact)
    trex_name <- file.path(outdir, paste0("trex_", i, ".txt"))
    write.table(contact, trex_name, row.names = FALSE, col.names = FALSE, quote = FALSE)
}

# Software Versions

```{r}
sessioninfo::package_info()
```

# Reference