---
title: "tRexCAP: A Divide and Conquer Strategy for Recapitulating Whole Genome 3D Structure Using Hi-C Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tRexCAP: A Divide and Conquer Strategy for Recapitulating Whole Genome 3D Structure Using Hi-C Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
The spatial organization of the genome is a critical determinant of cellular function, yet its complex architecture cannot be fully captured through linear sequencing alone. Hi-C technology overcomes this limitation by capturing chromosomal conformational information between loci. Translating the resulting 2D proximity matrices into physical 3D models enables the identification of spatial "neighborhoods," revealing how non-coding genetic variants exert regulatory influence over vast genomic distances.

tRex-CAP (Cut and Paste) is designed to reconstruct whole-genome, three-dimensional structures from Hi-C data using tREX (truncated Random effect EXpression) to generate its building blocks by employing a divide-and-conquer strategy to manage the complexity of genomic data.


Specifically, the tRex-CAP algorithm consists of two steps. The first step is `Cut` where the entire contact matrix is partitioned into n blocks (with overlaps), and tRex algorithm is run on each block to obtain its respective structure. The second step is `Paste` where all blocks are merged into one structure using isometric transformations to preserve the structure of each block while matching them at the overlapping loci.

We illustrate the usage of these functions in the following sections.

# Algorithm

## The Cut Function


### Arugments

The `Cut` function requries the following arguments:

* Contact: The matrix should be n by n and symmetric. The (i, j)-th entry of the matrix represent the count of interactions between loci i and j. The diagnoal elements of the matrix should be 0. An example contact matrix of n=1000 loci simulated with Poisson model with random effects can be loaded by calling `data(example_data)`.

```{r setup}
library(tRexCAP)
data(example_data)
dim(example_data)
example_data[1:5,1:5]
```

* block_size: The constatnt size of each block if the argument 'breaks' is not provided; otherwise, 'breaks' takes the precedence. (Note: This parameter is optional if `breaks` is provided. See `breaks` for details.)
* noverlap: Number of loci that overlap between neighboring blocks. The default value is 40.

Optional arugments include:

* bias: An n by 3 matrix. The 3 columns describe effective fragment information, GC content, and mappability for each locus, respectively. 

```{r bias}
data(example_bias)
head(example_bias)
```

* breaks: A vector that describes the indices of pre-specified cut points. If unspecified, equal block size cuts will be used according to the number specified in the `block_size' parameter. 

* CPU: Integer specifying the number of cores for parallel MCMC execution. Defaults to 1. 

* save_mcmc: Whether to save the posterior samples generated during the MCMC step. The default value is FALSE.

* mcmc, burn: The number of HMC iterations and burn-ins, respectively. The default values for both are 10,000.

* thinning: Thinning rate. For example, a thinning rate of 10 means every 10th HMC iteration is saved as posterior samples. The default value is 10.

* jump: The initial standard deviations for multiple parameters updated via Metropolis-Hastings. Refer to the in package for the default values.

* leapfrog.L, leapfrog.e: The initial number of leapfrog steps and intial epsilon for HMC. The default values are 10 for L and 0.001 for e.

* target.accept The target acceptance rate. The default values are 0.95.

* tuning_control: Tuning control for HMC. See below for details.

* method.type: The method for for reconstructing each block. One of tRex, tPAM and bn (Bayesian Nonparametric).

### Tuning Control
We do a five phase tuning. Phase 1 adapts the number of steps (L) and the step size (epsilon), phase 2 runs the chain, phase 3 adapts the step size (epsilon), phase 4 runs the chain, and phase 5 adapts the Metropolis-Hasting jump parameters.

The tuning control is a list comprising of `iter`, `block_size` and `max_cycle`. The `iter` parameter controls the number of iterations for each phase. Here, the `block_size` parameter controls how often acceptance rates are reported and how frequently hyperparameter adjustments occur according to parameter acceptance rates. The `max_cycle` parameter determines how many times the algorithm goes through the phases. Between cycles, the structure is reset but hyperparameter settings are retained. By default, the settings for `tuning_control` are retained. Below is an example of the `tuning_control` used for the sample data.

```{r controls}
control_trex <- list(
  iter = c(500, 500, 500, 500, 500),
  block_size = c(10, 10, 10, 10, 10),
  max_cycle = 1
)
```

### Output

The `Cut` function returns the sturcture of each block and other parameters intended for `Paste`.

## The Paste Function

The `Paste` function combines structures obtained in `Cut` step through rigid transformations.

### Arugments

* contact: The same contact matrix used in `Cut`.
* cutresult: The result from a Cut() call.
* CPU:  Integer specifying the number of cores for parallel MCMC execution. The default value is 1.

### Output
`Paste` returns an n by 3 matrix of the estimated coordinates.

## Example
Here we provide an example for tRex-CAP. For illustration simplicity, we use the first 100 loci in our example dataset with no bias matrix.

```{r, results = 'hide', message = FALSE}
example_data_short <- example_data[1:100, 1:100]
CutRes <- Cut(
            contact = example_data_short, 
            block_size = 25, 
            noverlap = 1, 
            CPU = 1,
            mcmc = 1000,
            burn = 1000,
            thinning = 1,
            target.accept = 0.95,
            tuning_control = control_trex
          )
```

On the 3303MHz AMD EPYC 7R32 CPU, without parallelization it took 14.7 seconds to execute the command above.

To paste the results, run the following script:
```{r, results = 'hide', message = FALSE}
PasteRes <- Paste(
              contact = example_data_short,  
              cutresult = CutRes, 
              CPU = 1
            )
```

The result is an n by 3 matrix.
```{r}
head(PasteRes)
```

On the 3303MHz AMD EPYC 7R32 CPU, without parallelization it took 23.8 seconds to execute the command above.

## The CutAndPaste Function
We also wrapped `Cut` and `Paste` into a single function call, `CutAndPaste`. The arguments for `CutAndPaste` are the same as `Cut`. It runs `Cut` and then `Paste` sequentially, and returns the n by 3 coordinates matrix.

```{r, eval=FALSE}
CAPRes <- CutAndPaste(
            contact = sim_hic,
            block_size = 12,
            noverlap = 1,
            CPU = 5)
```

# Visualization
Once we obtain the coordinate matrix, we can plot the 3D structure with `draw_struct`. `draw_struct` provides interactive 3D visualizations based on coordinates. 

# Application to the Lymphoblastoid Data
For users interested in reproducing our results with the lymphoblastoid data from lieberman:2009, we have provided the code below. Note that the parameter values are initially set to small values to allow for a quick test run. To fully reproduce the results reported in this study, users should use the significantly higher iteration counts specified in @pl:20161. We have processed the contact matrix to remove loci with zero interactions. The bias matrix is not used.

```{r, eval=FALSE}
data(Lieberman_breaks)
data(Lieberman_contact)

control_trex <- list(
  iter = c(500, 500, 500, 500, 500),
  block_size = c(10, 10, 10, 10, 10),
  max_cycle = 1
)

CutRes = Cut(contact = Lieberman_contact,
            breaks = Lieberman_breaks,
            noverlap = 1, 
            CPU = 1,
            mcmc = 100,
            burn = 100,
            thinning = 1,
            target.accept = 0.95,
            tuning_control = control_trex)

PasteRes = Paste(contact = Lieb_contact, 
                cutresult = CutRes, 
                CPU = 1)
                
```

Or, run `CutAndPaste` in one command:
```{r, eval = FALSE}
CAPRes = CutAndPaste(contact = Lieberman_breaks,
            breaks = Lieberman_contact,
            noverlap = 1,
            CPU = 1,
            mcmc = 1000,
            burn = 1000,
            thinning = 1,
            target.accept = 0.95,
            tuning_control = control_trex)
```

## Results
```{r}
head(PasteRes)
```

# Simulation
For users who are interested in reproducing the simulation results in @pl:20161, the following simulation function and script will lead to the reproduction of the same data sets. 

For a known structure `S`, we provided the `simulate_contact` function to simulate the data based on either Poisson or negative binomial distrubtion . Arguments for `simulate_contact` include `pi` for zero inflation percentage, `p` for overdispersion paramter for negative binomial distribution, and `random` for the presence of random effect. 

We used the first 500 and the last 500 loci from @lieberman:2009 as the structure. The coordinates whole genome is available from `data(lieberman)`.
```{r}
options(scipen = 999)
data(lieberman)
N <- nrow(lieberman)
# in our simulation we take the first 500 coordinates and last 500 coordinates of the leibermanS
S <- lieberman[c(1:500, (N - 499):N),]
head(S)
```

Below is the script we used for simulation. With the provided seed, the simulated data should be identical to those in @pl:20161. 
```{r, eval = FALSE}
require(VGAM)
require(reticulate)
require(splines)

# settings
family <- "poisson" # or nb for negative binomial 
pi <- 0 # zero-inflation parameter: options used in manuscript: 0, 0.05, 0.15, 0.25, 0.5
p <- 0.75 # overdispersion parameter: options used in manuscript: 0.5, 0.75, 0.9
random <- TRUE # random effect: if TRUE use random effect terms if FALSE do not use
nreps <- 1  # we set nreps to 100 in tRex-cap manuscript
seed = 2  # seed used in manuscript
cat("Simulating ", family, " with pi=", pi, ", p=", p, " random: ", random, " nreps = ", nreps, "\n", sep="")
set.seed(seed)

# build an informative output directory name
if (family == "poisson") {
  outdir <- paste0(
    "../siml_data/",
    family,
    "_pi_", pi,
    ifelse(random, "_ran", "_noran")
  )
} else {
  outdir <- paste0(
    "../siml_data/",
    family,
    "_pi_", pi,
    "_p_", p,
    ifelse(random, "_ran", "_noran")
  )
}

# create it if it doesn't exist
if (!dir.exists(outdir)) {
  dir.create(outdir, recursive = TRUE)
}


for (i in 1:nreps) {
    contact <- gen_contact(S, family = family, pi = pi, p = p, random = random)
    contact <- contact + t(contact)
    trex_name <- file.path(outdir, paste0("trex_", i, ".txt"))
    write.table(contact, trex_name, row.names = FALSE, col.names = FALSE, quote = FALSE)
}
```

# References