---
title: "tRexCAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tRexCAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Load data

An example contact matrix of n=43 loci can be loaded by calling `data(sim_hic)`. The matrix should be \eqn{n \times n} and symmetric, where the (i, j) entry of the matrix is the count of interactions between loci i and j. The diagnoal elements of the matrix should be 0.

The optional bias matrix should be of dimension \eqn{n \times 3}. The 3 columns describe effective fragment information, GC content, and mappability for each locus, respectively. 
```{r setup}
library(tRexCAP)
data(sim_hic)
data(sim_bias)
```

# Running Cut-and-Paste
Once the data is loaded, Cut-and-Paste algorithm can run separately by calling `Cut` and `Paste` in two separate steps.

```{r, eval=FALSE}
CutRes = Cut(contact = sim_hic, bias = sim_bias, block_size = 12, noverlap = 1, CPU = 5)
PasteRes = Paste(contact = sim_hic, cutresult = CutRes, CPU = 5)
```

Alternatively, it can be run in a single step by calling `CutAndPaste`
```{r, eval=FALSE}
CAPRes = CutAndPaste(contact = sim_hic, block_size = 12, noverlap = 1, CPU = 5)
```
